---
title: "How to succeed in this class"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")
```

## Syllabus

--

- Read the syllabus

--

- The syllabus is available on the edX platform

--

- There may be minor changes in syllabus from semester to semester;  
  in case of doubt, the version posted on edX is binding

---

## Course materials

--

- All slides and worksheets are also available at:  
  https://wilkelab.org/DSC385/

--

- For the curious, the source materials for the slides are here:  
  https://github.com/wilkelab/DSC385
  
--

- You can keep a local copy of the class site by downloading the `docs` folder and opening `index.html` in your browser:  
  https://github.com/wilkelab/DSC385/tree/main/docs

--

- The class uses extensive materials from my book, available here:  
  https://clauswilke.com/dataviz

---

## Homework and project submissions

--

- Make sure you actually submit your assignment

--

- Before submission, verify you have submitted the right files

--

- Once submitted, you can no longer resubmit

--

- You will need to peer grade before receiving your own grade

--

- EdX calculates the overall grade as the median of the first *n* grades

---

## Assignment format for submissions

We ask that you submit two files:

--

1. A pdf document representing your submission

--

2. A zip file containing the source files (.Rmd file, data files, etc) to reproduce the pdf

--

The pdf should be separate from the zip file, not included inside it, as that helps us to quickly look at an assignment.

--

Add some random letters or numbers to the file name: `project1_dlxdle.pdf`

--

**Do not** include your name or EID in the document or file name.


---

## In case of problems with submission, grading, etc.

--

Please do the following:

- Send email to: <a href="mailto:onlinedataexploration@austin.utexas.edu">onlinedataexploration@austin.utexas.edu</a>

--

- Include your edX user name

--

- Attach any assignment files if applicable

--

<br>

Please **do not:**
- post on Piazza
- send email directly to instructor or TA
- try to message on edX
- try to message on Canvas

---

## Peer grading

--

- Peer grading is an important component of this class

--

- The main purpose is for you to engage with the work of your peers

--

- There are two types of graded assignments: homeworks and projects

---

## Peer grading of homeworks

--

- Homeworks are primarily graded on effort:  
  Did the student make an honest attempt to complete the assignment?
  
--

- Homeworks **do not** have to match the example solution 1:1

--

- Please provide written comments justifying your grade

--

- Read the grading rubric!

--

- Project 1 is treated as a homework

---

## Peer grading of projects

--

- Projects have more complex rubrics. They differ somewhat among projects!

--

- Projects **do not** have to match the example solution 1:1

--

- Some projects do not have an example solution

--

- Please provide written comments justifying your grade

---

## Grading the reproducibility of a submission

--

- An assignment is reproducible if you can take the source files and reproduce the submitted pdf

--

- The following error is **not** a reproducibility problem:  
  `pdflatex: command not found`  
    
  You can resolve it by clicking the button to knit to HTML
  
--

- Assignments should only rely on the R packages listed as required in the syllabus

---

[//]: # "segment ends here"

---
class: center middle

## Making a reproducible example

---

## Making a reproducible example

--

- For questions on Piazza, please include a reproducible example (reprex)

--

- A reprex must include **all** code necessary to reproduce the problem

--

- A reprex should be as small as possible; delete what isn't needed

--

- It is always possible to create a good reprex, but sometimes it's a bit of work


---

## Include **all** code necessary to reproduce the problem

Example: Why does this code not run?

```{r eval = FALSE}
mtcars %>%
  ggplot(aes(disp, mpg)) + geom_point()
```

--

Error obtained:

```
## Error in mtcars %>% ggplot(aes(disp, mpg)) : 
##   could not find function "%>%"
```

---

## Include **all** code necessary to reproduce the problem

Complete code as run by student:

```{r eval = FALSE}
library(ggplot2)

mtcars %>%
  ggplot(aes(disp, mpg)) + geom_point()
```

```
## Error in mtcars %>% ggplot(aes(disp, mpg)) : 
##   could not find function "%>%"
```


---

## Include **all** code necessary to reproduce the problem

Complete code as run by student:

```{r eval = FALSE}
library(ggplot2)

mtcars %>%
  ggplot(aes(disp, mpg)) + geom_point()
```

```
## Error in mtcars %>% ggplot(aes(disp, mpg)) : 
##   could not find function "%>%"
```


Problem: Missing `library(tidyverse)`

---

## Include **all** code necessary to reproduce the problem

This code works:

```{r mtcars_working, eval = FALSE}
library(tidyverse)

mtcars %>%
  ggplot(aes(disp, mpg)) + geom_point()
```

.pull-left.width-35[
```{r mtcars_working-out, ref.label="mtcars_working", fig.width = 4, fig.asp = 0.75, echo=FALSE, warning=FALSE, message = FALSE, dev = "svg"}
```
]

--

.pull-right.width-60[
The problem is often not where you are looking for it.

]

---

## Make up data if necessary

A common concern: <br>

I cannot share my complete code. It would reveal my solution to the homework / assignment / etc.

--

<br>

Solution: <br>

Prepare a simplified example with made-up data

---

## Make up data if necessary

Example: Why are the data points not colored?

.tiny-font.pull-left[
```{r iris_broken, eval = FALSE}
library(tidyverse)

ggplot(iris) +
  aes(Sepal.Length, Sepal.Width) +
  aes(colr = Species) +
  geom_point(size = 2) +
  ggtitle("Measurements on iris flowers") +
  xlab("Sepal Length") +
  ylab("Sepal Width") +
  theme_bw()
```
]

.pull-right[
```{r iris_broken-out, ref.label="iris_broken", fig.width = 4, fig.asp = 0.75, echo=FALSE, warning=FALSE, message = FALSE, dev = "svg"}
```
]

---

## Make up data if necessary

Same example with made-up data:

.tiny-font.pull-left[
```{r madeup_broken, eval = FALSE}
library(tidyverse)

data <- tibble(
  x = 1:3,
  y = c(2, 1, 3),
  letter = c("A", "B", "C")
)

ggplot(data) +
  aes(x, y) + 
  aes(colr = letter) + # problem is here
  geom_point()
  # note: irrelevant code has been removed
```
]

.pull-right[
```{r madeup_broken-out, ref.label="madeup_broken", fig.width = 4, fig.asp = 0.75, echo=FALSE, warning=FALSE, message = FALSE, dev = "svg"}
```
]
---

## Make up data if necessary

Same example with made-up data:

.tiny-font.pull-left[
```{r madeup_fixed, eval = FALSE}
library(tidyverse)

data <- tibble(
  x = 1:3,
  y = c(2, 1, 3),
  letter = c("A", "B", "C")
)

ggplot(data) +
  aes(x, y) + 
  aes(color = letter) + # now it's fixed
  geom_point()
  # note: irrelevant code has been removed
```
]

.pull-right[
```{r madeup_fixed-out, ref.label="madeup_fixed", fig.width = 4.7, fig.asp = 0.618, echo=FALSE, warning=FALSE, message = FALSE, dev = "svg"}
```
]

---

## Closing thoughts about reproducible examples

--

- It can be a lot of effort to prepare a good, minimal example!

--

- You often end up solving the problem as you try to isolate it

--

- Use the **reprex** package to format your example: https://reprex.tidyverse.org/

--

- **reprex** works great with Piazza

--

- Copy and paste your actual code, not a screenshot; again, the **reprex** package helps

---

[//]: # "segment ends here"


---

## Further reading

- All slides and worksheets: [wilkelab.org/DSC385](https://wilkelab.org/DSC385/)
- Github repo of class materials: [https://github.com/wilkelab/DSC385](https://github.com/wilkelab/DSC385)
- **reprex** package: [https://reprex.tidyverse.org/](https://reprex.tidyverse.org/)
